---
title: "Pharmacy Operations Analytics"
author: "Kush Manek"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---
```{r}

library(tidyverse)
library(lubridate)
library(janitor)

set.seed(123)

### 1. technicians.csv (20,000 rows)
technicians <- tibble(
  technician_id = sprintf("T%03d", 1:20000),
  name = paste0("Tech_", sprintf("T%03d", 1:20000)),
  experience_years = sample(1:10, 20000, replace = TRUE),
  shift_type = sample(c("Day","Evening","Mixed"), 20000, replace = TRUE)
)

write_csv(technicians, "technicians.csv")

### 2. patients.csv (20,000 rows)
cities <- c("Victoria","Saanich","Langford","Sidney","Nanaimo")

patients <- tibble(
  patient_id = sprintf("P%05d", 1:20000),
  age = sample(18:95, 20000, replace = TRUE),
  sex = sample(c("F","M","Other"), 20000, replace = TRUE),
  chronic_conditions_count = sample(0:4, 20000, replace = TRUE),
  insurance_type = sample(c("Public","Private","None"), 20000, replace = TRUE),
  city = sample(cities, 20000, replace = TRUE)
)

write_csv(patients, "patients.csv")

### 3. inventory.csv (20,000 rows)
drug_classes <- c("Antibiotic","Antidepressant","Analgesic","Antihypertensive","Antidiabetic")
forms <- c("Tablet","Capsule","Syrup","Injection")

inventory <- tibble(
  drug_id = sprintf("D%03d", 1:20000),
  drug_name = paste0("STAT_D", sprintf("%03d", 1:20000)),
  drug_class = sample(drug_classes, 20000, replace = TRUE),
  form = sample(forms, 20000, replace = TRUE),
  strength = paste0(sample(c(5, 10, 20, 50, 100), 20000, replace = TRUE), "mg"),
  stock_on_hand = sample(0:500, 20000, replace = TRUE),
  reorder_level = sample(20:100, 20000, replace = TRUE),
  lead_time_days = sample(1:30, 20000, replace = TRUE),
  unit_cost = round(runif(20000, 5, 200), 2),
  expiry_date = sample(seq.Date(as.Date("2024-01-01"), as.Date("2026-12-31"), by="day"), 20000, replace = TRUE)
)

write_csv(inventory, "inventory.csv")

### 4. prescriptions.csv (20,000 rows)
dates_2024 <- seq.POSIXt(
  as.POSIXct("2024-01-01 08:00:00"),
  as.POSIXct("2024-12-31 20:00:00"),
  by = "min"
)

prescriptions <- tibble(
  prescription_id = sprintf("RX%06d", 1:20000),
  patient_id = sample(patients$patient_id, 20000, replace = TRUE),
  technician_id = sample(technicians$technician_id, 20000, replace = TRUE),
  drug_id = sample(inventory$drug_id, 20000, replace = TRUE),
  datetime = sample(dates_2024, 20000, replace = TRUE),
  quantity = sample(c(10, 15, 20, 30, 60, 90), 20000, replace = TRUE),
  days_supply = sample(c(7, 14, 30, 60, 90), 20000, replace = TRUE),
  refill = sample(0:1, 20000, replace = TRUE),
  insurance_type = sample(c("Public","Private","None"), 20000, replace = TRUE),
  wait_time_min = round(rnorm(20000, mean = 12, sd = 4), 1),
  status = sample(c("Completed","Rejected","Cancelled"), 20000, replace = TRUE, prob = c(0.9,0.05,0.05))
)

write_csv(prescriptions, "prescriptions.csv")

### 5. otc_sales.csv (20,000 rows)
otc_categories <- c("Vitamins","Pain Relief","Cold/Flu","Digestive Health","Allergy")
otc_products <- c("Ibuprofen","Acetaminophen","Multivitamin","Vitamin D","Omega-3","Loratadine","Probiotic")

otc_sales <- tibble(
  sale_id = sprintf("S%06d", 1:20000),
  datetime = sample(dates_2024, 20000, replace = TRUE),
  product_category = sample(otc_categories, 20000, replace = TRUE),
  product_name = sample(otc_products, 20000, replace = TRUE),
  quantity = sample(1:5, 20000, replace = TRUE),
  unit_price = round(runif(20000, 5, 50), 2),
  revenue = quantity * unit_price,
  promotion_flag = sample(0:1, 20000, replace = TRUE),
  patient_id = sample(c(patients$patient_id, rep(NA, 5000)), 20000, replace = TRUE)
)

write_csv(otc_sales, "otc_sales.csv")
```



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)
library(scales)
library(glue)
library(knitr)
library(kableExtra)
library(forecast)
library(timetk)
library(tsibble)
library(lme4)
library(ggplot2)

theme_set(theme_minimal())

safe_read <- function(path) {
  if(file.exists(path)) read_csv(path) %>% clean_names()
  else stop(glue("File not found: {path}"))
}
```

# 1. Introduction

This R Markdown provides a reproducible end-to-end analytics workflow for synthetic pharmacy operations data covering the year 2024.

Key goals:
- Prescription volume and peak times  
- Technician performance & wait times  
- Inventory risk & stockout simulation  
- OTC sales behavior and promotion lift  
- Forecasting demand and operational recommendations  


# 2. Data loading and basic checks

Place these CSVs in the same folder:
- prescriptions.csv  
- inventory.csv  
- technicians.csv  
- patients.csv  
- otc_sales.csv  

```{r load-data}
prescriptions <- safe_read("prescriptions.csv")
inventory     <- safe_read("inventory.csv")
technicians   <- safe_read("technicians.csv")
patients      <- safe_read("patients.csv")
otc_sales     <- safe_read("otc_sales.csv")

glimpse(prescriptions)
glimpse(inventory)
glimpse(technicians)
glimpse(patients)
glimpse(otc_sales)

tibble(dataset = c("prescriptions","inventory","technicians","patients","otc_sales"),
       n = c(nrow(prescriptions), nrow(inventory), nrow(technicians), nrow(patients), nrow(otc_sales))) %>%
  kable() %>% kable_styling(full_width = FALSE)
```


# 3. Data preparation and enrichment

```{r prep}
prescriptions <- prescriptions %>%
  mutate(
    datetime = ymd_hms(datetime),
    date = as_date(datetime),
    hour = hour(datetime),
    wday = wday(datetime, label = TRUE)
  )


otc_sales <- otc_sales %>%
  mutate(
    datetime = ymd_hms(datetime),
    date = as_date(datetime),
    hour = hour(datetime),
    wday = wday(datetime, label = TRUE)
  )


prescriptions <- prescriptions %>% mutate(patient_id = as.character(patient_id),
                                         technician_id = as.character(technician_id),
                                         drug_id = as.character(drug_id))

patients <- patients %>% mutate(patient_id = as.character(patient_id))
technicians <- technicians %>% mutate(technician_id = as.character(technician_id))
inventory <- inventory %>% mutate(drug_id = as.character(drug_id))
otc_sales <- otc_sales %>% mutate(patient_id = as.character(patient_id))

rx_full <- prescriptions %>%
  left_join(patients, by = "patient_id") %>%
  left_join(technicians, by = "technician_id") %>%
  left_join(inventory, by = "drug_id")

missing_tech <- rx_full %>% filter(is.na(technician_id)) %>% nrow()
missing_drug <- rx_full %>% filter(is.na(drug_id) | is.na(drug_name)) %>% nrow()

cat(glue("Missing technician assignments: {missing_tech}\nMissing drug lookups: {missing_drug}\n"))
```


# 4. Exploratory Data Analysis

## 4.1 Prescription volume (daily & hourly)

```{r rx-volume}
daily_rx <- rx_full %>% count(date)

ggplot(daily_rx, aes(date, n)) +
  geom_line() +
  labs(title = "Daily Prescription Volume (2024)",
       y = "Prescriptions per day") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b")

hourly_rx <- rx_full %>% count(hour)

ggplot(hourly_rx, aes(hour, n)) +
  geom_col() +
  labs(title = "Prescription Volume by Hour",
       x = "Hour", y = "Count") +
  scale_x_continuous(breaks = 0:23)
```


## 4.2 Weekday × hour heatmap

```{r heatmap}
heat_df <- rx_full %>%
  group_by(wday, hour) %>%
  summarise(n = n(), .groups = "drop")

ggplot(heat_df, aes(hour, fct_rev(wday), fill = n)) +
  geom_tile() +
  labs(title = "Heatmap: Prescriptions by Weekday and Hour",
       x = "Hour", y = "Weekday")
```


# 5. Wait Time Analysis & Technician Performance

## 5.1 Wait time distribution

```{r wait-dist}
ggplot(rx_full, aes(wait_time_min)) +
  geom_histogram(bins = 40) +
  labs(title = "Distribution of Wait Times",
       x = "Wait time (minutes)")

avg_wait_hour <- rx_full %>%
  group_by(hour) %>%
  summarise(avg_wait = mean(wait_time_min, na.rm = TRUE))

ggplot(avg_wait_hour, aes(hour, avg_wait)) +
  geom_line() + geom_point() +
  labs(title = "Average Wait Time by Hour",
       x = "Hour", y = "Avg wait (minutes)")
```


## 5.2 Technician performance

```{r tech-perf}
tech_wait <- rx_full %>%
  group_by(technician_id, name) %>%
  summarise(
    avg_wait = mean(wait_time_min, na.rm = TRUE),
    rx_count = n(),
    avg_experience = mean(experience_years, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(tech_wait, aes(reorder(name, avg_wait), avg_wait)) +
  geom_col() +
  coord_flip() +
  labs(title = "Average Wait Time by Technician",
       x = "Technician", y = "Avg wait (min)")

tech_wait %>% arrange(avg_wait) %>% head(5) %>% kable() %>% kable_styling()
tech_wait %>% arrange(desc(avg_wait)) %>% head(5) %>% kable() %>% kable_styling()
```


# 6. Mixed-effects model for technician impact

```{r tech-model}

library(lme4)

use_ins <- "insurance_type" %in% names(rx_full)

model_data <- rx_full %>%
  select(wait_time_min, technician_id, experience_years, hour,
         !!(if (use_ins) sym("insurance_type") else NULL)) %>%
  drop_na()

formula <- if (use_ins) {
  wait_time_min ~ insurance_type + experience_years + hour + (1 | technician_id)
} else {
  wait_time_min ~ experience_years + hour + (1 | technician_id)
}

m1 <- lmer(formula, data = model_data)

summary(m1)

# Random intercepts (technician effects)
ranef_te <- ranef(m1)$technician_id %>%
  as.data.frame() %>%
  tibble::rownames_to_column("technician_id")

names(ranef_te) <- c("technician_id", "rand_intercept")

ranef_te %>%
  arrange(desc(rand_intercept)) %>%
  head() %>%
  kable() %>%
  kable_styling()

```


# 7. Inventory Risk & Stock Depletion Simulation

```{r inventory-risk}
inventory <- inventory %>%
  mutate(stock_gap = stock_on_hand - reorder_level,
         at_risk = stock_on_hand <= reorder_level)

inventory %>% count(at_risk)

inventory_at_risk <- inventory %>%
  filter(at_risk) %>%
  arrange(stock_on_hand)

inventory_at_risk %>% head(20) %>% kable() %>% kable_styling()
```


## Stockout projection

```{r inventory-simulation}
daily_demand <- prescriptions %>%
  count(drug_id, date) %>%
  group_by(drug_id) %>%
  summarise(avg_daily = mean(n))

inv_sim <- inventory %>%
  left_join(daily_demand, by = "drug_id") %>%
  mutate(avg_daily = if_else(is.na(avg_daily), 0.5, avg_daily),
         days_until_reorder = (stock_on_hand - reorder_level) / avg_daily,
         projected_days_to_stockout = stock_on_hand / avg_daily,
         reorder_by_date = Sys.Date() + ceiling(days_until_reorder),
         expected_stockout_date = Sys.Date() + ceiling(projected_days_to_stockout),
         needs_reorder_now = stock_on_hand <= reorder_level)

inv_sim %>%
  select(drug_id, drug_name, stock_on_hand, reorder_level,
         avg_daily, needs_reorder_now, reorder_by_date, expected_stockout_date) %>%
  head(20) %>% kable() %>% kable_styling()
```


# 8. Forecasting Daily Prescription Volume (30 days)

```{r forecast-30d, fig.height=4, fig.width=8}

# Ensure date column is valid
if (!"date" %in% names(rx_full))
  stop("rx_full is missing `date` column — earlier parsing failed.")

daily_rx <- rx_full %>%
  filter(!is.na(date)) %>%     # remove bad dates
  count(date) %>% 
  arrange(date)

# Check for empty dataset
if (nrow(daily_rx) == 0) {
  stop("No valid date values found in prescriptions — cannot run forecast.")
}

# Fill missing dates (important for time series)
full_dates <- seq(min(daily_rx$date), max(daily_rx$date), by = "day")

daily_rx_full <- tibble(
  date = full_dates
) %>%
  left_join(daily_rx, by = "date") %>%
  mutate(n = ifelse(is.na(n), 0, n))

# Build ts object safely
rx_ts <- ts(
  daily_rx_full$n,
  frequency = 365
)

# Fit model
fit <- auto.arima(rx_ts)

# Forecast 30 days
fcast <- forecast(fit, h = 30)

# Plot
autoplot(fcast) +
  labs(
    title = "30-Day Forecast: Daily Prescription Volume",
    x = "Date",
    y = "Prescriptions"
  )

# Forecast dataframe
fcast_df <- tibble(
  date = seq(max(daily_rx_full$date) + 1, by = "day", length.out = 30),
  point_forecast = as.numeric(fcast$mean),
  lo80 = as.numeric(fcast$lower[,1]),
  hi80 = as.numeric(fcast$upper[,1]),
  lo95 = as.numeric(fcast$lower[,2]),
  hi95 = as.numeric(fcast$upper[,2])
)

fcast_df %>% head(10) %>% kable() %>% kable_styling(full_width = FALSE)


```

# 9. OTC Sales & Promotion Lift Analysis

```{r otc-analysis}
promo_summary <- otc_sales %>%
  mutate(promotion_flag = as.integer(promotion_flag)) %>%
  group_by(product_category, promotion_flag) %>%
  summarise(
    total_revenue = sum(revenue),
    total_qty = sum(quantity),
    n_sales = n(),
    .groups = "drop"
  ) %>%
  pivot_wider(names_from = promotion_flag,
              values_from = c(total_revenue, total_qty, n_sales),
              names_prefix = "promo_") %>%
  mutate(lift_revenue = (total_revenue_promo_1 /
                           pmax(total_revenue_promo_0, 1)) - 1) %>%
  arrange(desc(lift_revenue))

promo_summary %>% head(15) %>% kable() %>% kable_styling()

otc_sales %>% count(wday) %>%
  ggplot(aes(wday, n)) +
  geom_col() +
  labs(title = "OTC Sales by Day of Week",
       x = "Day", y = "Sales Count")
```


# 10. Business Insights & Recommendations

- **Peak hours** show strong lunchtime and evening surges → increase staffing at 11–2pm and 4–7pm.  
- **Technician performance** varies meaningfully; experience reduces wait time → schedule senior techs during peak.  
- **Inventory**: several SKUs require immediate reorder → short lead-time drugs prioritized.  
- **OTC promotions**: categories with highest lift should get recurring targeted promotions.  
- **Forecast** indicates expected surge periods → adjust technician scheduling and inventory planning.  


# 11. Power BI Integration Notes

- Create relationships:  
  prescriptions → patients  
  prescriptions → technicians  
  prescriptions → inventory  
  otc_sales → patients (nullable)

Suggested pages:
- RX Overview  
- Wait Time Diagnostics  
- Inventory & Stockouts  
- OTC Promotions & Revenue Lift  


# 12. Appendix

```{r session-info}
sessionInfo()
```


